An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-03-27 RESPONSE.txt

General comment: Two of the reviews seem to ask (more or less
directly) what the contribution of the paper is.  We admit that the
introduction may not have provided a sufficiently clear overview over
this work, since many logics are introduced, but in the following we
try to briefly outline this (and we will of course later improve the
introduction).

In previous work [23,24] we presented a BDD-based monitoring algorithm
for:

  PFLTL : Past-time First-order LTL

as defined on page 9. This logic has limitations in expressiveness as
shown with two examples on page 10. As main contribution, we
consequently present an extension of this logic, called:

  EPFLTL: Extended Past-time First-ordrer LTL

See page 11 equation (5). The extension coniststs of adding rules to
temporal properties.  Example properties (not expresssible in PFLTL)
are shown in Figure 2 page 16.  This logic is strictly more expressive
than PFLTL by allowing e.g. Wolper's in-every-even-state property and
generally transitive closures over data in the trace to be
expressed. The algorithm is presented on page 14.  The corresponding
tool has been extended as well (and is availbale open source).

The paper is structured in such a way that it gradually builds up this
result in steps, with some intermediate results, first making the
argument for the propositional LTL case in Sections 2 and 3, and then
for the first-order LTL case in Sections 4 and 5, as follows.

For the propositional case:

- LTL is introduced and its limitation to past time operartors PLTL
  (known results)

- QLTL (and its past time version QPLTL) is LTL (PLTL) with (so-called
  dynamic) quantification over states. This is a known extension
  described by Wopler in [21]. This extends the expressiveness of LTL
  to allow formulation of the property: "every even state satisfies
  p", not expressible in LTL. However it is not obvious how to design
  an algorithm for efficiently monitoring QLTL properties. Therefore
  we alternatively propose (and this is the main theme in the paper):

- ELTL (and its past time version EPLTL) that instead extends LTL
  (PLTL) with rules, achieving the same expressivness as QLTL (QPLTL)
  but with a well defined monitoring algorithm. This is a new result.

For the first-order case we repeat the above story, ending with the final
contribution:

- FLTL and specifically PFLTL are introduced. The logic PFLTL and its
  monitoring algorithm was introduced in previous work by us [23,24].

- QFLTL (and its past time version QPFLTL) is introduced, showing how
  to extend FLTL with quantification over states as in the
  propositional case. This is a new theoretic result. However, as in
  the propositional case, it is not clear how to construct an
  efficient monitoring algorithm. This therefore leads us to the final
  result and main contribution:

- EFLTL and specifically the past time version EPFLTL are introduced,
  and a monitoring algorithm for EPFLTL is presented.

----------------------- REVIEW 1 ---------------------

> The authors claim that these extensions are useful for runtime
> verification and present some minimal experimental results from
> running a BDD-based implementation of the extensions.

As for experiments with BDDs, we have performed many experiments, they
are mentioned in our past papers, and are publically available with
our tool. But due to anonymous reviewing, we cannot reveal it. We only
here presented experiments to show the strength of the extension of
the logic.

> I am not quite on board with the motivation of the paper. In runtime
> verification, the truth values of formulas are determined on a single
> trace, hence, in particular, LTL and CTL are equivalent, and the
> verification is hence polynomial. It is also quite standard in
> model-checking to augment the intermediate states on the path with the
> truth values of subformulas (this is how CTL model checking is done,
> and hence also LTL for a single trace). Wouldn't it address at least
> some of the concerns of the authors (that is, which subformulas are
> satisfied in which states)?

*** TODO ***

We do not quite understand this comment.

> Also, I see no reference to PSL, which is the IEEE standard for
> property specification languages, or to ForSpec, which is Intel's
> specification language. I am wondering whether these would already
> provide the necessary functionality.

We should definitely cite the excellent well known work on
PSL/ForSpec. However, PSL and other standards are limited in the way
they handle data, and are basically strong propositional logics. They
are equivalent to logics that we cited, e.g. regular expressions and
2nd order monadic logic (this is not a logic with data!). We
introduced the propositional extension with rules (ELTL/EPLTL) to
demonstrate how simple (not more expressive than regular expressions)
our extension is, and as an easy to read motivation for a related
extension for first-order LTL (EFLTL/EPFLTL).  The subject of our
paper is not about expressiveness of propositional LTL, but a
practical extension to first-order LTL. As a side effect, the
propositional introduction also makes a new nice style of expressive
LTL properties.

> Adding quantification is solvable, in general, with a QBF solver, and
> in some cases even without it. In the Intel's paper about enhanced
> vacuity from the beginning of 2000's, they studied the problem of
> model-checking a quantified formula and demonstrated that for the
> types of formulas they constructed it is no harder than
> model-checking.

We assume you are referring to the paper:

  Enhanced Vacuity Detection in Linear Temporal Logic
  R. Armoni, L. Fix, A. Flaisher, O. Grumberg, N. Piterman, A. Tiemeyer, M. Vardi
  CAV 2003

The main difference is the first-order nature of our logic, in
contrast to the logic presented in that paper, which is propositional
(and by the way limited in where quantifiers can appear). However,
thanks for the reference, we will explore this work.

> Finally, I might be wrong, but it seems to me that BDDs are no longer
> widely used in model-checking, replaced by SAT and SMT solvers. I am
> wondering whether the implementation with BDDs is interesting enough.

The main problem studied in this paper is that of a linear temporal
logic with data, i.e., (past) first order linear temporal logic. For
this reason, a representation that can efficiently handle relations
over (basically unbounded) data is important. The BDD representation
is used to represent such big relations, where the BDDs hold
enumerations of the seen data values rather than that (hashed) data
values themselves. Consequently, we can use efficient standard Boolean
operators on BDDs to implement intersection, union and complementation
of sets (since the sets are only semi infinite, there is a special
encoding value for "all values not seen"). This is a surprisingly
efficient representation which works with first-order LTL.  BDDs are
nowdays less frequently used for model checking, but (a) this is not
model checking but runtime verification (b) they are used for the
specific and ingenious way of representing relations, and (c) SAT
solvers are not needed here. We are aware of (and will cite) work on
using SMT for runtime verification:

  Monitoring Modulo Theories
  Normann Decker, Martin Leucker, and Daniel Thoma

but it is not particularly efficient.

----------------------- REVIEW 2 ---------------------

> While the paper explores a potentially useful abstraction for runtime
> verification, it was hard to fully understand the motivation for this
> work. Some ways that the work could be strengthened are listed below.

Sure, see also leading comment above.

> Thm. 1 and Lemma 2 show that the introduced logics (ELTL and EPLTL)
> have the same expressive power as Q(P)LTL. The authors should make the
> benefit of introducing these new logics more clear.
> 
> Thm. 2 shows that
> QPFLTL is more expressive than EPFLTL, but the authors do not provide
> any complexity results on the decidability of these logics - for what
> benefit are we trading the expressivity?

As explained in the general introduction of this response, the main
reason for moving from 'Q' to 'E', e.g. from QPFLTL to EPFLTL, is that
we can show how to construct an efficient monitoring algorithm for 'E'
(e.g. EPFLTL), whereas we do not see how to do this for more theoretic 'Q'
(e.g. QPFLTL). That is the main message of the paper.

> Thm 2 also opens the question
> of whether EQPFLTL makes sense and if so how it relates to the other
permutations of (E/Q/F/PLTL).

There is no need to extend QPFLTL with rules since the rule extension 'E'
is an easier monitorable *alternative* to the dynamic quantification 'Q'.

> The limitations of existing runtime verification systems for data
> intensive systems could be made more clear through an evaluation that
> compares the use of EFPLTL for runtime verification with the
> alternative existing approaches. At the moment the evaluation only
> presents benchmarks for the introduced tool, but does not situate this
> evaluation in the context of existing runtime verification tools.

That is a fair comment.

> Furthermore, the evaluation includes a 'spawning' trace where analysis
> time appears to scale exponentially in the size of the trace. The
> authors mention this is possibly due to the increasingly large amount
> of data that must be stored as the trace grows. As the system is
> intended to be used for runtime verification, what practical
> consideration must be made given the exponentially scaling with
> respect to the lifecycle of the task being monitored?

It is certainly true that dealing with data introduces the problem
that monitoring becomes trace length dependent.  To tackle this
problem, the original version of the tool performs a clever form of
garbage collection, described in [24], where data no longer needed are
dropped from the BDDs.  In the here (artificially created) property
and trace, there is indeed a problem if too many threads are spawned.
We have not in this work studied how to limit monitoring to become
less trace length dependent. We consider this as an orthogonal,
although very important, line of work in the field of runtime
verification. Only little work has been done in this direction so far.

> In addition, a more thorough analysis of memory usage should be
> included in the evaluation. No information is provided on the system
> that ran these evaluation results. Do these benchmarks' memory usage
> exceed the available RAM and trigger disk operations? As the temporal
> relations of (P)FLTL require increasingly large hash tables to save
> their evaluations, what impact will this have on memory limited
> systems? This is especially relevant for the many use cases of runtime
> verification on embedded systems.

Indeed, memory analysis should be included. This information was not
directly visible when running our scripts (not a good excuse,
admitted). Memory is of course critical. As mentioned above, we do
perform garbage collection on unused values as described in [24]. For
example for the property

  forall x . p(x) -> !q(x) S r(x)

(meaning: forall x . if p(x) then in the past r(x) and since then no
q(x)), the monitor will drop any value x for which we in the past saw
an r(x) and since then saw! a q(x). But the general problem of
monitoring embedded systems is challenging. Not all problems can be
solved at once.

> Another issue is that it was hard to identify the paper's
> contribution. The section on 'Runtime verification algorithm for
> PFLTL' has already been described in [24]. To handle EPFLTL, Sec 5
> introduces a single 'rename' function which seems a relatively small
> delta with respect to [24]. The authors point us back to section 3
> (mixed evaluation order) for further contributions with respect to
> [24] to handle EPFLTL, but it is then not clear why Section 3 is
> separate from Section 5.

See general introduction to this response for a more general outline
of the work. The handling of EPFLTL, compared to PFLTL, is not trivial
and was quite difficult to implement. We do not consider it a
relatively small delta, although the theory perhaps is small (and
elegant). Concering the question why Section 3 is separate from
Section 5: as outlined in the introductory text above, Section 3 deals
with the propositional case whereas Section 5 deals with the
first-order case.

> The paper should better position its contributions wrt. related work.
> Throughout the paper, the authors should make it more obvious to the
> reader that only logics prefixed with an E are new logics introduced
> in this work (if I understood the paper correctly).

Well, QFLTL and QPFLTL are new (first-order LTL extended with dynamic
quantification).  But yes, the comment is largely correct. Our
introduction did not clarify the structure well enough.

> Additionally, the use of predicates to encode information about the
> prefix of the state (system inputs seen so far), and thereby allowing
> arbitrary data over infinite domains, was explored in 'Synthesizing
> Functional Reactive Programs' by Finkbeiner et. al
> https://arxiv.org/abs/1712.00246.

*** TODO ***

> Finally, the paper contains typos that create problematic ambiguities.
> As an example, the top of page 5 notes that the logic PQLTL is
> obtained by restricting QPLTL to past modalities. From context, it
> appears this should read as QLTL, not QPLTL.

Indeed.

> Furthermore, by page 6
> the authors are proving properties of QPLTL, but PQLTL is no longer
> referenced. Although from context the reader can infer that QPLTL and
> PQLTL these should be taken as the same logic (except on page 5 when
> QPLTL refers to QLTL), these inferences should not be left to the
> reader. I also believe that there is an issue with Theorem 2 - when
> comparing the expressive power of EPFLTL and QFPLTL, the statement
> instead reads as comparing EPFLTL and QFLTL.

Indeed to all observations here, we should have used/should use latex
macros for this!

----------------------- REVIEW 3 ---------------------

> The experimental results seem
> preliminary: apart from the fact that the examples are made up, it's
> unclear how the collection of traces was generated.

The artificial traces were generated by a trace generating program
parameterized with various constants, indicating how many events of
various kinds should be generated, and the patterns with which they
are generated. The objectieve was to generate as many data values in
the trace as possible, while at the same time also cause repetition of
patterns.

> It would be more
> convincing to see this applied to more realistic examples (esp the
> transitive closure property surely has many applications).  However,
> the tool is apparently available publicly (links hidden due to blind
> review).

Yes, it would be better with more realistic examples. The tool is
publicly available.

------------------------------------------------------

### 2019-03-28 RESPONSE-V2.txt

Contribution: in previous work [23,24] we presented a BDD-based
monitoring algorithm for PFLTL: Past-time First-order LTL. This logic
has limitations in expressiveness exemplified on page 10. As main
contribution, we consequently present an extension of this logic (page
11):

  EPFLTL: Extended Past-time First-ordrer LTL

The extension adds rules to temporal properties. Example properties
(not expressible in PFLTL) are shown (Figure 2).

We build this result in steps, making the argument initially for
propositional LTL (Sections 2,3) and then for first-order LTL
(Sections 4,5).

For each case (propositional and first-order) we present three
logic-classes.  Let X stand for LTL/PLTL (propositional case) or
FLTL/PFLTL (first-order case):

  core form       :  X
  quantified form : QX (extends expressiveness but no monitoring algorithm known)
  extended form   : EX (extends expressiveness with shown monitoring algorithm)

----------[REVIEW-1]----------

> and present some _minimal_ experimental results

We have presented many experiments in previous papers, which are
publicly available with our tool. We here added experiments to show
the strength of the extension.

> verification is hence polynomial ...

For unbounded data and unbounded sequences, the size of information
that needs to be stored can grow unboundedly. Our BDD representation,
however, shows impressive results, handling large traces.

> Also, I see no reference to PSL, ... ForSpec, ...

We will cite this work. However, PSL is limited in the way it handles
data, and is basically a strong propositional logic. It is equivalent
to some propositional logics we cite.  We introduced the propositional
extension with rules to demonstrate how simple (not more expressive!)
our extension is, that a simple extension to a classical RV algorithm
handles it, and to motivate the related extension for first-order LTL.

> In the Intel's paper about enhanced
> vacuity from the beginning of 2000's, they studied the problem of
> model-checking a quantified formula

There are two types of quantifiers: propositional quantifiers, as
QLTL, as dealt with in the paper you mention, and first-order
quantification (in our logics named with letter 'F'). Our contribution
is w.r.t. handling the first-order version, so the work mentioned is
not connected.

> ... BDDs are no longer widely used in model-checking, replaced by SAT and SMT ..

The main problem studied in handling RV for first-order LTL with data
is the representation of relations over data, which are accumulated
throughout the execution. In our BDD encoding of relations, we can use
efficient Boolean BDD operators to implement intersection, union and
complementation of relations. This is surprisingly efficient allowing
handling of very large traces with lots of data.

----------[REVIEW-2]----------

> have the same expressive power as Q(P)LTL. The authors should make the
> benefit of introducing these new logics more clear.
> Thm. 2 shows that
> QPFLTL is more expressive than EPFLTL...for what benefit are we trading the expressivity?

The main reason for moving from 'Q' to 'E', e.g. from QPFLTL to
EPFLTL, is that we can show how to construct an efficient monitoring
algorithm for 'E' (e.g. EPFLTL), whereas we do not see how to do this
for 'Q' (e.g. QPFLTL). We do not believe there is a reasonable
algorithm for the 'Q' version and the example in the proof gives some
insight of what is involved.

> the evaluation only presents benchmarks for the introduced tool

In [23] we compared with MonPoly, which is closest, since it includes
FPLTL.

> time appears to scale exponentially in the size of the trace.

The BDD representation of relations showed on many examples (open
source) to be quite compact. The original version of the tool also
performs a clever form of garbage collection, described in [24].
Indeed, BDDs are not a magic fix-all tool, but we obtain impressive
results.

> a more thorough analysis of memory usage should be included

Memory analysis is missing, agreed.

> The paper should better position its contributions wrt. related work.

Yes, although see also introduction to this response above.

> Additionally, the use of predicates ... was explored in 'Synthesizing Functional Reactive Programs'.

We were not aware of this paper (it appears as an archived paper).  It
seems that the TSL assignments only assign one value to a
variable. Our approach maintains sets of such values, needed for
monitoring. But we do not claim to have introduced predicates to
encode information about the prefix of the state, we refer to paper
[4].

----------------------- REVIEW 3 ---------------------

> unclear how the collection of traces was generated.

The artificial traces were generated by a trace generating program
parameterized with various constants, indicating how many events of
various kinds should be generated, and the patterns with which they
are generated. The objective was to generate as many data values in
the trace as possible, while at the same time also cause repetition of
patterns.

> more convincing to see this applied to more realistic examples

Agreed!

### 2019-04-18 Response/RESPONSE-V1.txt

> The paper should better position its contributions wrt. related work.
> Throughout the paper, the authors should make it more obvious to the
> reader that only logics prefixed with an E are new logics introduced
> in this work (if I understood the paper correctly).

Well, QFLTL and QPFLTL are new (first-order LTL extended with dynamic
quantification).  But yes, the comment is largely correct. Our
introduction did not clarify the structure well enough.

> Additionally, the use of predicates to encode information about the
> prefix of the state (system inputs seen so far), and thereby allowing
> arbitrary data over infinite domains, was explored in 'Synthesizing
> Functional Reactive Programs' by Finkbeiner et. al
> https://arxiv.org/abs/1712.00246.

The paper referred to indeed introduces a concept of assignments in
the temporal logic TSL.  An interesting read (nice paper)! we were not
aware of this work. There are some essential differences though.

First of all, assignments in TSL are to simple variables that along
one trace only can hold one value at a time. This in contrast to our
solution, where rules (and quantified variables in general) can keep
track of multiple values (sets of values, they are relations in
general), which is necessary for expressing the kind of properties we
are interested in. E.g. it is as far as we understand (we may be
wrong) not possible in TSL to express the following property, assuming
that file identifiers range over the infinite domain of strings:

  "whenever a file x is opened, file x should not be zip'ed
   until file x has been closed"

Assume that numerous files can be open at the same time, an unknown
number up front.  If in TSL one uses a variable that stores a single
file name this will not work.  To handle such a property, a variable
storing file identifiers would need to be able to store all files
opened so far and not yet closed. This of course can be programmed by
letting a variable store a set of file names. However, this is like
programming, and it becomes more complicated with more complicated
properties. Another way of phrasing this is that TSL is closer to the
concept of extended state machines, where there is one state machine
with one state of data values that is updated in addition to the
control states changing. In our framework we can say that we have a
state machine for each value x.

Second, TSL does not support the transitive closure of relations,
whereas our logic does.  This is non-trivial as well.

Finally, TSL is a future time logic, whereas our logic is a past time
logic. There is a substantial difference in how algorithms in these
two different spaces work.

### 2019-04-18 implementation.tex
started editing small things according to reviews

\section{Implementation}

@@

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The properties are 
%
The ASCII version of the logic here uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

@@

The second property, \idsl{telemetry2}, expresses the same 
property as \idsl{telemetry1}, but in this case using two 
rules, more closely reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
\idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
(the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
\idsl{closed(x)} was true in the previous state and there is
no \idsl{toggle(x)} event.
The third alternative states that we in the previous state were
in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} 
event. Similarly for the \idsl{open(x)} rule.

>>>>

\section{Implementation}
\label{implement}

@@

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input ASCII format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. 
%
The ASCII version of the logic uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

@@

The second property, \idsl{telemetry2}, expresses the same 
property as \idsl{telemetry1}, but in this case using two 
rules, more closely reflecting how we would model this using a state machine with two states for each channel \idsl{x}: 
\idsl{closed(x)} and \idsl{open(x)}. The rule \idsl{closed(x)} is defined as a disjunction between three alternatives. The first states that this predicate is true if we are in the initial state 
(the only state where \idsl{@true} is false), and there is no \idsl{toggle(x)} event. The next alternative states that
\idsl{closed(x)} was true in the previous state and there is
no \idsl{toggle(x)} event.
The third alternative states that in the previous state we were
in the \idsl{open(x)} state and we observe a \idsl{toggle(x)} 
event. Similarly for the \idsl{open(x)} rule.

### 2019-04-18 rule-ltl.tex
started editing small things according to reviews

Linear Temporal Logic (LTL) is a common specification formalism for reactive and concurrent systems. It is often used in e.g. model checking and runtime verification. Another formalism that is used for the same purpose is finite automata, often over infinite words. This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power. In fact, model checking of an LTL specification
is usually performed by first translating the specification into a B\"{u}chi automaton. The automata formalisms are more expressive than LTL, with a classical example By Wolper~\cite{Wolper} showing that
it is not possible to express in LTL that every even state in
the sequence satisfies $p$. This has motivated extending LTL
in various ways.
We are mainly motivated by runtime verification of executions {\em with data}, for which a first-order temporal logic is
appropriate. There, too, we show that expressiveness can be extended.

@@

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the unexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
We present an extension where auxiliary relations, not appearing in the execution, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding full quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
provides the full expressive power of, say, B\"{u}chi automata.

@@

Restricting QPLTL to the past modalities, one obtain the logic PQLTL. PQLTL has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in PQLTL as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.

%Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
%PQLTL property into a deterministic
%automaton.

@@

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$.
%A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_j$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $B$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j$ as a {\em rule}. In text, rules
 will be separated from each other by commas. 
\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

@@

\begin{lemma} \label{fourone}
Let $\varphi$ be an ELTL formula
over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
a set of rules $a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$. 
Let $\sigma$ be a model with states over the $B$. 
Then there is at most one
model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $

@@

\subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
formula into a deterministic finite automaton, where one distinguished states, $s$, in a sink state.
Then the sequence of states to be checked forms the input for this automaton, and violation is
announced if $s$ is reached.
QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

@@

We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We allow the following formula:
\begin{equation}
\label{EFLTL}
\ \ \ \ \ \psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \mathrm{\ \ \ \     such\ that,}
\end{equation}
\begin{enumerate}
\item $\psi$, the {\em statement}, is an FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formulas with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
type as $x_j$ and the second one is, as usual,
a natural number that is omitted in the temporal formulas. 
An auxiliary relations $r_j$ can appear within $\psi$. They can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
scope
of a previous-time operator $\ominus$. 
\end{enumerate}

@@

%We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
%and QLTL (PQLTL, respectively) have the same
%expressive power. We study now the relationship between
%EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).

\begin{theorem}
\label{theo2}
The expressive power of EPFLTL is strictly weaker than that of QFLTL.
\end{theorem}

@@

We use a combinatorial argument to show by contradiction that one cannot express this property
in EPFLTL. The reason is that every prefix of an a model for an EPFLTL property is extended uniquely with
auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in
the model, the auxiliary relations and the
assignments satisfying the subformulas. The size of
each such relation is bounded by ${\cal O} (  m ^ N )$
where $m$ is the number of values appearing in the
prefix, and $N$ is the number of parameters of the relations.
However, the number of different data sets over $m$ values is
$2^m$. This means that with large enough number of different values, each EPFLTL formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not. The semantics of EPFLTL implies that extending
two models with the same summary in the same way would have the same truth value. So, we can extend
the two models with a data set that appears in one of the
prefixes but not  in the other, and the EPFLTL property will not be able to distinguish between them. \qed

>>>>

Linear Temporal Logic (LTL) is a common specification formalism for reactive and concurrent systems. It is often used in e.g. model checking and runtime verification. Another formalism that is used for the same purpose is finite automata, often over infinite words. This includes B\"{u}chi, Rabin, Street, Muller and Parity automata~\cite{Thomas}, all having the same expressive power. In fact, model checking of an LTL specification
is usually performed by first translating the specification into a B\"{u}chi automaton. The automata formalisms are more expressive than LTL, with a classical example by Wolper~\cite{Wolper} showing that
it is not possible to express in LTL that every even state in
the sequence satisfies $p$. This has motivated extending LTL
in various ways.
We are mainly motivated by runtime verification of executions {\em with data}, for which a first-order temporal logic is
appropriate. There, too, we show that expressiveness can be extended.

@@

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the inexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
We present an extension where auxiliary relations, not appearing in the execution, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding full quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
provides the full expressive power of, say, B\"{u}chi automata.

@@

Restricting QLTL to the past modalities, one obtains the logic QPLTL. QPLTL has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in QPLTL as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.

%Note that since finite automata over finite sequences can be determinized, it is always sufficient to translate a
%QPLTL property into a deterministic
%automaton.

@@

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$.
%A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_j$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $A$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j$ as a {\em rule}. In text, rules
 will be separated from each other by commas. 
\begin{tabbing}
\ \ \ $\sigma \models \eta$ \=  if there exists $\sigma'$, where
$\sigma' |_A = \sigma$ such that \\
 \> $\sigma' \models ( \psi  \wedge   \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j)) $
\end{tabbing}

@@

\begin{lemma} \label{fourone}
Let $\varphi$ be an ELTL formula
over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
a set of rules $a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$. 
Let $\sigma$ be a model with states over $B$. 
Then there is at most one
model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $

@@

\subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
formula into a deterministic finite automaton, where one distinguished state, $s$, is a sink state.
Then the sequence of states to be checked forms the input for this automaton, and violation is
announced if $s$ is reached.
QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

@@

We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We allow the following formula:
\begin{equation}
\label{EFLTL}
\ \ \ \ \ \psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} \mathrm{\ \ \ \     such\ that,}
\end{equation}
\begin{enumerate}
\item $\psi$, the {\em statement}, is an FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formulas with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
type as $x_j$ and the second one is, as usual,
a natural number that is omitted in the temporal formulas. 
An auxiliary relation $r_j$ can appear within $\psi$. They can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
scope
of a previous-time operator $\ominus$. 
\end{enumerate}

@@

%We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
%and QLTL (QPLTL, respectively) have the same
%expressive power. We study now the relationship between
%EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).

\begin{theorem}
\label{theo2}
The expressive power of EPFLTL is strictly weaker than that of QPFLTL.
\end{theorem}

@@

We use a combinatorial argument to show by contradiction that one cannot express this property
in EPFLTL. The reason is that every prefix of a model for an EPFLTL property is extended uniquely with
auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in
the model, the auxiliary relations and the
assignments satisfying the subformulas. The size of
each such relation is bounded by ${\cal O} (  m ^ N )$
where $m$ is the number of values appearing in the
prefix, and $N$ is the number of parameters of the relations.
However, the number of different data sets over $m$ values is
$2^m$. This means that with large enough number of different values, each EPFLTL formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not. The semantics of EPFLTL implies that extending
two models with the same summary in the same way would have the same truth value. So, we can extend
the two models with a data set that appears in one of the
prefixes but not  in the other, and the EPFLTL property will not be able to distinguish between them. \qed
